Synthesizing Quotient Lenses

### 2017-10-17 papers/QuotientLensSynth/main.tex
another draft

\usepackage{amsmath, amssymb, verbatim, enumerate, 
graphicx, centernot, tikz, array, tikz-cd, extarrows, cleveref,
mathrsfs, mathtools, bussproofs, stmaryrd, enumitem, stackengine}

@@

  However, quotient lenses are often difficult to program.  The programmer has
  to give a specification for the data on each side of the lens and a separate
  specification for the equivalence relations defined on the data. The
  programmer must also manually implement a lens program that maps between the
  specifications while respecting the equivalence relations---this is itself a
  challenging task thanks to the restrictions on how lenses can be combined
  while still satisfying the lens laws.

@@

\section{Introduction}
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

\noindent stored in the \bibtex{} format to their equivalent representation 
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A Conway, J. H.
%F conway
\end{verbatim}

@@

Bidirectional programs have also been called {\em lenses}. Formally, Foster et
al~\cite{foster} defined a lens $\ell$ that maps between an ``abstract" {\em
source} $S$ and a ``concrete" {\em view} $V$ to be a triple of functions
\begin{align*}
\ell.get &: S \longrightarrow V\\
\ell.put &: V \longrightarrow S \longrightarrow S\\
\ell.create &: S \longrightarrow V
\end{align*}

\noindent satisfying the equations
\begin{align*}
\ell.\lput \; (\ell.\get \; s) \; s &= s\\
\ell.\get \; (\ell.\lput \; v \; s) &= v\\
\ell.\lput \; (\ell.\create \; v) &= v\\
\end{align*}

@@

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in
reality are not part of the data. Foster {\em et al.\relax} therefore proposed
\textit{quotient lenses} ~\cite{quotient} as a solution to this problem
Quotient lenses, which are implemented as a refinement of the bidirectional
string processing language Boomerang, enable a programmer to write lenses which
transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

\sam{Go over quotient lens laws?}

Unfortunately, quotient lenses are sometimes difficult to program in Boomerang,
firstly because the programmer has to give a specification for the data as well
as give a separate specification for the equivalence relations defined on the
data. For example, suppose that the programmer wants to program the \bibtex{} to
EndNote transformation in such a way that the program allows for any of two
representations of the authors name (e.g. ``Conway, J. H.'' and ``Conway
J. H''\bcp{``J. H. Conway''??}), 
ignores the order in which the author and title fields occur in each record,
and allows for arbitrary amounts of whitespace in between the author and title
fields. These rules allow for these two \bibtex{} citation records to be
considered the same,\bcp{I thought
  we were going to allow newlines within titles and authors...?}

\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}

@Book{conway,
    Title = {Regular Algebra and Finite Machines},

    Author = {Conway J. H.},
}
\end{verbatim}

with the EndNote records having a similar equivalence relation defined on them.

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map each element of a regular language to a
canonical representation of that element. Therefore, in order for the programmer
to give a specification for the \bibtex{} records, then the programmer
has to write a regular expression which matches all possible representations of
\bibtex{} records, and then program a set of canonizers which does the job of
quotienting the data. This often makes the code verbose and error-prone.

\sam{TODO: PUT BOOMERANG CANONIZERS HERE!}

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which enables the programmer to explicitly define a broad
class of equivalence relations directly on regular languages using a compact
and convenient notation, and hence negates the need to specify the data and the
equivalence relation on the data separately. For example, a QRE which
specifies the equivalence relation on the \bibtex{} citation records is as
follows:

\begin{verbatim}
let bib_author  = 
"author = {" . squash bib_space_names -> bib_comma_names using 
(get (synth bib_space_names <=> bib_comma_names)) . "}"

let bib_ignore_wsp =  project ("," . [ \n\t\r]* ) -> ",\n"

let bib_perm : canonizer = perm (label, bib_author, bib_title) with
bib_ignore_wsp

let bibtex  = "@book{" . bib_perm . "}"
\end{verbatim}

\sam{Should we have a special construct like squash-synth(R, R') in addition to
squash(R, R', l), or maybe even just have only squash-synth(R, R')?}

The first expression specifies the author field as
having one of two representations (\verb!bib_space_names! and
\verb!bib_comma_names!), with the \verb!bib_comma_names! being the
canonical representation and the expression

\noindent \verb!(get (synth bib_space_names <=> bib_comma_names))! giving the
function which converts names described by \verb!bib_space_names! to their canonical
form described  by \verb!bib_comma_names!.

The second expression specifies the regular language which is a comma followed
by arbitrary occurences of whitespace characters quotiented by the equivalence
relation which makes all members of this language equivalent, with the canonical
representation being a comma followed by a newline character.

The third expression specifies that the label, title and author fields
are can occur in any order, with each of these fields separated by a comma and
arbitrary amounts of whitespace. The canonical representaion is that in which
the label, author and title fields occur in that order, separated by a comma and
a newline.

However, even though QREs enable the programmer to specify the data as well as
the equivalence relation on the data in the same expression, the programmer
still has to manually program the underlying lens which maps the equivalence
classes of the \bibtex{} records to the equivalence classes of the EndNote
records

\sam{TODO: PUT BIJECTIVE LENS HERE!},

\noindent task which is itself challenging and erro-prone.

Therefore, to address this challenge we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs. Our
key insight is to reduce the problem of synthesizing QRE lenses to the problem
of synthesizing {\em bijective lenses}, a problem which we solved in out prior
work ~\cite{popl18}.

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with canonizers at the ends. That is, a QRE lens uses the
source QRE to compute a unique representative for the data modulo the
equivalence relation defined by the QRE before applying a bijective lens to
this representative in the forward direction, and operates similarly same in the
backward direction, but using the view QRE.

\sam{TODO: PUT LENS WITH CANONIZERS ON THE EDGE DIAGRAM HERE!}

One potential pitfall with QRE lenses is that the composition of QRE lenses may
not have the normal form of a bijective lens with canonizers on the ends:

\sam{TODO: PUT COMPOSITION DIAGRAM HERE!}

Our main technical contribution in this paper is a proof that QRE lenses are
closed under left quotienting, right quotienting, composition, and regular
combinators. This proof will allow justify our approach for synthesizing
quotient lenses so that the programmer can simply type

$$\mathit{synth} \; \verb!bibtex! \; \verb!endnote!$$

to derive a quotient lens that does the \bibtex{} to EndNote transformation.
\bcp{What about examples??}

@@

For example, suppose that the programmer wishes to write a regular expression
which describes \bibtex{} citation records such as:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Each record may be represented in many different ways. Firstly, the author's
name can have the last name followed by a comma, and then the
first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. For example, each of these
\bibtex{} records is equivalent to the other:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = {Conway J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Secondly, the author and title fields may occur in either order:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Title = {Regular Algebra and Finite Machines},
  Author = {Conway, J. H.},
}
\end{verbatim}
Thirdly, the title and author records can be separated by arbitrary amounts of
whitespace characters. For example, if the title field occurs first, then the
author field may occur not on the line immediately below the title field, but
instead on the second line below the title field:
\begin{verbatim}
@Book {conway,
  Title = {Regular Algebra and Finite Machines},

  Author = {Conway, J. H.},
}
\end{verbatim}
In the example above, a \bibtex{} record may have different
representations for three different reasons. The first reason
is that the author's name can have the last name followed by a comma and then
the first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. The $\squash{R}{R'}{f}$ QRE
enables the programmer to choose between the representations described by $R$
and $R$', provided that they can provide a function $f$ that maps data stored in
the first format to data stored in the second format. In this example we have
the two regular expressions

\begin{align*}
\mathtt{COMMA\_NAME} &= \mathtt{NAME} \cdot ``," \cdot \mathtt{NAME} \cdot ``
\quad " \cdot \mathtt{NAME}\\
\mathtt{SPACE\_NAME} &= \mathtt{NAME} \cdot ``," \cdot \mathtt{NAME} \cdot ``
\quad " \cdot \mathtt{NAME}\\
\end{align*}

and the bijective lens 
$$ \ell = (\mathit{copy} \; \mathtt{COMMA\_NAME}) \cdot (\mathit{ins} \; ``,")
\cdot (\mathit{del} \; `` \quad") \cdot (\mathit{copy} \; \mathtt{COMMA\_NAME})
\cdot (\mathit{copy} \; \mathtt{COMMA\_NAME})$$
therefore the QRE
$$\mathtt{AUTHOR}
=
``\text{Author
= \{}" \cdot \squash{\mathtt{COMMA\_NAME}}{\mathtt{SPACE\_NAME}}{\ell.get}
\cdot ``\}"$$ enables the programmer to express the fact that the two different
representations of the author's name are equivalent, with the
$\mathtt{COMMA\_NAME}$ representation chosen as the canonical representation,
and the function $\ell.get$ giving the ``canonizing'' function which converts
data in the $\mathtt{SPACE\_NAME}$ format to its $\mathtt{COMMA\_NAME}$
representation. The $\squash{R}{R'}{f}$ has the added advantage that if the
transformation from $R$ to $R'$ can be expressed by a bijective lens,
then the programmer may attempt to \textit{synthesize} the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:
$$\mathtt{AUTHOR}
=
``\text{Author
= \{}" \cdot \squash{\mathtt{COMMA\_NAME}}{\mathtt{SPACE\_NAME}}{\mathit{synth}
\; \mathtt{COMMA\_NAME} \; \mathtt{SPACE\_NAME}} \cdot ``\}"$$

The second reason that a record may have two different representations is that
the author and title fields may occur in a different order, and the third
reason is that the author and title fields may have arbitrary occurences of
whitespace characters in between them. The $R \mapsto s$ and $\perm{q_1,
\ldots, q_n}{q}$ QREs enable the programmer to handle each of these cases:
$$\perm{\re{AUTHOR}, \re{TITLE}}{((``," \cdot \re{WHITESPACE}^*) \mapsto \string
``, \backslash n")}$$
This QRE enables the programmer to consider any permutation of the author and
title fields interspersed with a comma and arbitrary whitespace as equivalent.
The $((``," \cdot \re{WHITESPACE}^*) \mapsto \string ``, \backslash n")$
subexpression is itself a QRE that denotes that all string $s$ which consists
of a comma $``,"$ followed by any number of whitespace should be considered
equivalent, with the canonical representative being the string $\string ``,
\backslash n"$. The canonical representation of \bibtex{} records is that in
which the author field occurs the title field with a comma then a newline
character separating the two.

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} \; record is named
$\mathtt{RECORD}$, then the QRE for matching two records is $\mathtt{RECORD}
\cdot \mathtt{RECORD}$, the QRE for matching one or two records is $\mathtt{RECORD}
\sep (\mathtt{RECORD} \cdot \mathtt{RECORD})$ and the QRE for matching any
number of QRE records in $\mathtt{RECORD}^*$.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each of the
the other QREs $q$, there exist regular expressions $R, R'$ and a surjective,
idempotent function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that
$q$ $\normalize{R}{R'}{f}$ each denote the same equivalence relation on the
same regular language. $\normalize{R}{R'}{f}$ enables the programmer to
provide a canonizing function from $\mathcal{L}(R)$ to $\mathcal{L}(R')$,
with the equivalence class on $\mathcal{L}(R)$ being the equivalence relation
defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and only if $f(r) =
f(r'))$. The canonizing function is required to be surjective and idempotent.
The $\normalize{R}{R'}{f}$ combinator therefore enables the programmer to express
equivalence relations which are difficult to express, or which cannot be
expressed using the other combinators.

@@

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined. For example if $\mathtt{RECORD}$ is the QRE for a \bibtex
record in our \bibtex{} to and $W(\mathtt{ENDNOTE} \cdot \mathtt{ENDNOTE})$ is not
unambiguous, that is there is a string $s$ such that $s = s_1 \cdot s_2 =
{s_1}' \cdot {s_2}'$ with $s_1, s_2, {s_1}', {s_2}' \in W(\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE})$, then the canonizer $f$ for $\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE}$ may not be well defined since we are not guaranteed that
$f(s_1) \cdot f(s_2) = f({s_1}') \cdot f({s_2}')$. We also require that the
regular combinators applied to the kernels are unambiguous since the underlying
lenses will end up operating on the kernels, and these lenses impose the same
restrictions for similar reasons.

>>>>

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine}

@@

  However, quotient lenses are often difficult to program.  The programmer
  often has to give a specification for the data on each side of the lens and a
  separate specification for the equivalence relations defined on the data. The
  programmer must also manually implement a lens program that maps between the
  specifications while respecting the equivalence relations---this is itself a
  challenging task thanks to the restrictions on how lenses can be combined
  while still satisfying the lens laws.

@@

\section{Introduction}
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

\noindent stored in the \bibtex{} format to their equivalent representation 
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A J. H. Conway
%F conway
\end{verbatim}

@@

Bidirectional programs have also been called {\em lenses}. Formally, Foster et
al~\cite{foster} defined a lens $\ell : S \Leftrightarrow V$ that maps between
an ``abstract" {\em source} $S$ and a ``concrete" {\em view} $V$ to be a triple
of functions $(\ell.\get, \; \ell.\lput, \; \ell.\create)$ satisfying the
following equations:
\begin{align*}
\ell.get &: S \longrightarrow V & \ell.\lput \; (\ell.\get \; s) \; s &= s\\
\ell.put &: V \longrightarrow S \longrightarrow S & \ell.\get \; (\ell.\lput \;
v \; s) &= v\\
\ell.create &: S \longrightarrow V & \ell.\lput \; (\ell.\create \; v) &= v
\end{align*}

@@

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in
reality are not part of the data. Foster {\em et al.\relax} therefore proposed
\textit{quotient lenses} ~\cite{quotient} as a solution to this problem.
Quotient lenses, which are implemented as a refinement of the bidirectional
string processing language Boomerang, enable a programmer to write lenses which
transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

Formally, given equivalence relations $\sim_S$ and $\sim_V$ defined on the
source $S$ and the view $V$ respectively, a quotient lens $q : S/{\sim_S}
\Leftrightarrow V/{\sim_V}$ is a triple of functions $(\ell.\get, \; \ell.\lput,
\; \ell.\create)$ which satisfy the lens laws up to equivalence:
\begin{align*}
\ell.get &: S \longrightarrow V & \ell.\lput \; (\ell.\get \; s) \; s &\sim_S
s\\
\ell.put &: V \longrightarrow S \longrightarrow S & \ell.\get \; (\ell.\lput \;
v \; s) &\sim_V v\\
\ell.create &: S \longrightarrow V & \ell.\lput \; (\ell.\create \; v) &\sim_V v
\end{align*}
Moreover, each of the $\get$, $\lput$ and $\create$ functions must respect
$\sim_S$ and $\sim_V$. That is, if $s \sim_S s'$ then $\get \; s \sim_V
\get \; s'$, if $s \sim_S s'$ and $v \sim_V v'$ then $\lput \; v \; s \sim_S
\lput \; v' \; s'$ and if $v \sim_V v'$, then $\create \; v \sim_S \create \;
v'$.

In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map members of a regular language to their canonical
representation modulo an equivalence relation. Therefore, in order to give a
specification for the data, the programmer often has to write a
regular expression which matches all possible representations of the data, and
then provide a set of canonizers which does the job of quotienting the data.

For example, suppose that the programmer wants to
derive the \bibtex{} to EndNote transformation in such a way that the program
allows for whitespace to occur arbitrarily between the words in the author and
title fields, ignores the order in which the author and title fields occur in
each record, and allows for arbitrary amounts of whitespace in between the
author and title fields. These rules allow for these two \bibtex{} citation
records to be considered the same,

\bcp{I thought we were going to allow newlines within titles and authors...?}

\sam{I wanted to illustrate how to use the squash QRE which is why I changed my
mind on that, but on second thought since the problem of quotienting away
whitespace at will is going to be more familiar to the reader I changed it back
to the original plan. I will explain squash in the QRE section.}
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines},
}

@Book{conway,

    Title = {Regular Algebra 
            and Finite Machines},

    Author = "J. H. Conway",
}
\end{verbatim}

\noindent with the EndNote records having a similar equivalence relation defined
on them.

The code sample below shows how the programmer might specify these equivalence
relations on the \bibtex{} records in Boomerang using canonizers:

\begin{verbatim}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))

let wsp : regexp = [ \n\t\r]+

let wsp_to_space : canonizer = project wsp " "

let valid_name : canonizer = 
name | (name . wsp_to_space . (name .wsp_to_space)* .name)

let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "

let bib_names : canonizer =

valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)* . valid_name)

let bib_author : canonizer = "author = \"" . bib_names . "\""

let label : regexp = [a-zA-Z0-9]+

let title : canonizer = 
word | (word . wsp_to_space . (word . wsp_to_space)* . word)

let bib_title : canonizer = "title = {" . title . "}"

let wsp_to_comma : canonizer = project ("," . wsp) ",\n"

let first_perm : canonizer = bib_author . wsp_to_comma . bib_title

let second_perm : canonizer = bib_title . wsp_to_comma . bib_author

let bibtex : canonizer = 
"@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{verbatim}

Using the in-built \verb!canonizer_of_lens! function which extracts a canonizer
from a lens, it is easy to define a an equivalence relation that sends all
members of a regular language to a canonical representative. For example
the expression 
\begin{verbatim}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{verbatim}

\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space. However, a problem arises when trying to specify
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

Specifically, it becomes necessary to specify each of the permutations
seperately, since Boomerang's in-built permutation combinator does not account
for separators in between the regular expressions being permuted:

\begin{verbatim}
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title

let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
\end{verbatim}

One possible solution is to permute the alternative canonizers

\noindent \verb!bib_author. wsp_to_comma! and \verb!bib_title . wsp_to_comma!,
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers. In any case, the
regular expression
\begin{verbatim}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{verbatim}
is not equivalent to,
\begin{verbatim}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{verbatim}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \verb!bib_author! and \verb!bib_title!
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \verb!uncanonized_type! and \verb!canonized_type! which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

In any case, all of these alternatives lead to verbose code that is
difficult to understand. Moreover, if the \bibtex{} records had 3 or more
fields or even nested permutations, then the programmer would quickly find
themselves in a bind with these workarounds.

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which express a broad class of equivalence relations
directly on regular languages using a compact and convenient notation.

With QREs we can specify \bibtex{} records as follows:

\begin{verbatim}
let wsp_to_space : canonizer = project [ \n\t\r]+ -> " "

let valid_name : canonizer = 
name | (name . wsp_to_space . (name . wsp_to_space)* . name)

let wsp_to_and : canonizer = 
project ([ \n\t\r]+ . "and" . [ \n\t\r]+) -> " and "

let bib_names : canonizer =

valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)* . valid_name)

let bib_author : canonizer = "author = \"" . bib_names . "\""

let label : regexp = [a-zA-Z0-9]+

let title : canonizer = word | (word . wsp_to_space . (word . wsp_to_space)* .
word)

let bib_title : canonizer = "title = {" . title . "}"

let bib_perm  : canonizer = perm (label, bib_author, bib_title) with
(project ("," . [ \n\t\r]* ) -> ",\n")

let bibtex : canonizer = "@book{" . bib_perm . "}"
\end{verbatim}

The expression \verb!project [ \n\t\r]+ -> " "! is a QRE which matches any
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

\begin{center}
\begin{verbatim}
perm (label, bib_author, bib_title) with (project ("," . [\n\t\r]* ) -> ",\n")
\end{verbatim}
\end{center}

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \verb!perm! QRE is defined at the level of canonizers which allows for
easy nesting of permutations. Similarly, it is easy to specify any number of
fields within the \verb!perm! QRE.

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, he/she
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below illustrates:
\begin{verbatim}
let preamble : (lens in "@book{" <=> "%0 Book\n%F ") = 
    del "@book{" . ins "%0 Book\n%F "

let author_lens : (lens in (",\n" . bib_author) <=> ("\n" . end_author)) = 
    del ",\nauthor = \""
    . ins "\n%A "
    . valid_name
    . (del " and "
         . ins "\n"
         . (ins "%A " . valid_name . del " and " . ins "\n")*  . ins "%A "
         . valid_name
           || "")

let title_lens : 
(lens in ("\",\n" . bib_title . "},\n}") <=> ("\n" . end_title)) =
del "\",\ntitle = {" . ins "\n%T " . title . del "},\n}"

let bib_to_end : (lens in bib_tex <=> end_note) =
preamble . label . author_lens . title_lens
\end{verbatim}

This lens transforms the \bibtex{} data to the EndNote (and vice versa) data by
matching the delimiters and separators in a bijective manner. For example,
the \verb!preamble! lens replaces the \bibtex{} preamble \verb!@book! with the
EndNote preamble \verb!%0 Book\n%F ! in the \bibtex{record}. The programmer
can ill afford to make a mistake with any one of the small but vital details
of the transformations since even a one character mistake might render the lens
incorrect. This makes lens programming tedious and error-prone.

Therefore, to address this challenge, we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs.
Our key insight is to reduce the problem of synthesizing QRE lenses to the
problem of synthesizing {\em bijective lenses}, a particular subset of the
lenses that can be expressed in Boomerang. This is a problem which we solved in
our prior work ~\cite{popl18}.

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with ``canonizers at the edges'' as in
Figure~\ref{fig:attheedges}. In the forward direction, a QRE lens $q$ uses the
source QRE $c$ to compute a unique representative for the data modulo the
equivalence relation defined by $c$ and then applies the $\get$ function of a
bijective lens $\ell$ to this representative. In the backward direction, $q$
operates similarly, but using the view QRE $c'$ and the $\lput$ function of
$\ell$.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{attheedges}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
  screenshot from Nate's thesis. Are we allowed to use it?}}
  \label{fig:attheedges}
\end{figure}

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Our main technical contribution in this paper is a
proof that QRE lenses are closed under left quotienting, right quotienting,
composition, and regular combinators. This proof will allow justify our
approach for synthesizing quotient lenses so that the programmer can simply
type

\begin{verbatim}
let bib_to_end : (lens in bibtex <=> endnot) 
= synth bibtex <=> endnote  using  {(bib_example, end_example)}
\end{verbatim}
\noindent to synthesize a quotient lens that maps between \verb!bibtex! and
\verb!endnote! QREs, and sends the equivalence class of \verb!bib_bexample! to
the equivalence class of \verb!end_example!.

@@

We have already encountered the \verb!project R -> s! QRE which sends all
elements of the language described by the regular expression $R$, and the

\noindent \verb!perm (q_1, ..., q_n) with q! QRE which sends all permutations of
the QREs $q_1, \ldots, q_n$ separated by the QRE $q$ to the canonical
QRE $q_1 \cdot q \cdot \ldots \cdot q \cdot q_n$.

Now consider a variant of the \bibtex{} to EndNote example transformation,
where, the only quotienting occurs in the author field, where the author's name
can have the last name followed by a comma and then the rest of the name, or
remain as before where the names are separated by a space.

For example, these two records should be considered equivalent:
\begin{verbatim}
@Book {conway,
  Author = "Conway, J. H.",
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

The equivalence described is one in which there is a bijection between 
the regular expressions \verb!comma_name! and \verb!space_name! given by,
\begin{verbatim}
let comma_name : regexp = name | (name . ", " . name . (" " . name)*)

let space_name : regexp = name | (name . " " . (name . wsp_to_space)* . name)

\end{verbatim}

\noindent with the lens \verb!comma_to_space! mapping between them given by 
\begin{verbatim}
let comma_to_space : (lens in space_name <=> comma_name) =
lens_swap 
(name . del " " . (ins " " . name . del " ")* ) (name . ins ", ") || name 
\end{verbatim}

The $\squash{R}{R'}{f}$ QRE is designed to make it easy to express such
equivalences easily. Specifically, given regular expressions $R$ and $R'$, and
a function $f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$, the
$\squash{R}{R'}{f}$ QRE expresses the equivalence relation on $\mathcal{L}(R \;
R')$ which sends each element $r \in \mathcal{L}(R)$ to the canonical element
$f(r) \in \mathcal{L}(R')$, and sends each element $r' \in \mathcal{L}(R')$ to
itself:

\begin{verbatim}
let squash_author : canonizer =  
squash space_name -> comma_name using (get comma_to_space)

let author : canonizer = "author = \"" . squash_author . "\""
\end{verbatim}

In this example, the \verb!squash_author! QRE expresses the fact that the
\verb!space_name! and \verb!comma_name! representations of the authors' name
are equivalent, with the \verb!comma_name! representation chosen as the
canonical representation, and the function \verb!comma_to_space.get! giving
the canonizing function which converts data in the \verb!space_name! format
to its \verb!comma_name! representation.

The $\squash{R}{R'}{f}$ has the added advantage that if the transformation from
$R$ to $R'$ can be expressed by a bijective lens that is consistent with a set
of examples, then the programmer can synthesize the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:

\begin{verbatim}
let comma_to_space : (lens in space_name <=> comma_name) = 
synth space_name <=> comma_name using {("J. H. Conway", "Conway, J. H.")}
\end{verbatim}

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex{} record is named
\verb!bibtex!, then the QRE for matching two records is 
\verb!bibtex . bibtex!, the QRE for matching one or two records is
\verb! bibtex | (bibtex . bibtex)! and the QRE for matching any number of QRE
records \verb!bibtex*!

The $\normalize{R}{R'}{f}$ expresses the equivalence relation which sends each
element in $\mathcal{L}(R)$ to its canonical representation $f(r) \in
\mathcal{L}(R')$. The equivalence relation defined by
$\normalize{R}{R'}{f}$ is thus the equivalence relation on
$\mathcal{L}(R)$ defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and
only if $f(r) = f(r'))$. The canonizing function $f$ is required to be
surjective and idempotent.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each QREs
$q$, there exist regular expressions $R, R'$ and a surjective, idempotent
function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that $q$ and
$\normalize{R}{R'}{f}$ each define the same equivalence relation on
$\mathcal{L}(R)$, with $\mathcal{L}(R')$ forming a complete set of
representatives for this equivalence relation. The $\normalize{R}{R'}{f}$
combinator is included so as to express equivalence relations which are
difficult to express, or which cannot be expressed, using the other
combinators. However, checking that the canonizing function $f$ is
surjective and idempotent is undecidable, thus we do not use the
$\normalize{R}{R'}{f}$ in practice. On the other hand, the
$\normalize{R}{R'}{f}$ combinator gives a sufficient condition for potential
QREs to be ``well-behaved'', particularly with respect to our strategy for
synthesizing quotient lenses which we shall introduce in Section~\ref{synth}.

While QREs express a broad class of equivalence relations on regular languages,
they do not express all the equivalence relations that can be expressed in
Boomerang, nor of course all the equivalence relations which cannot be
expressed using Boomerang canonizers. For instance, Boomerang canonizers can
map a regular language described by $R$ to a regular language described by
$S$, with $\mathcal{L}(S) \not \subseteq \mathcal{L}(R)$, whereas each QRE $q$
canonizes a regular language described by $R$ to regular subset
$\mathcal{L}(R') \subseteq \mathcal{L}(R)$ described by $R'$.

@@

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined.

For example consider the QRE

\begin{verbatim}
let ambiguous : canonizer = "a"* . (project "a"* -> "a")
\end{verbatim}

\noindent The behaviour of \verb!ambiguous! is not-well defined since the string
\verb!"aaa"! can be canonized to any of \verb!"a", "aa", "aaa"! depending on how
\verb!"aaa"! is parsed. We also require that the regular combinators applied to
the kernels are unambiguous since the underlying lenses will end up operating
on the kernels, and these lenses impose the same restrictions for similar
reasons.
